---
title: "Worsts Bad Weather Events Cause Enormous Losts in Lives and Properties."
author: "herchu1"
output: html_document
---

## Synopsis

Although the citizens are aware of the damages that severe bad weather
events produce due to the general public media, the total quantities of losts
in human life and properties throughout the years it's not well known.
We analized the weather events that causes the worst damages. We compared
the events identifying which provokes more lives lost against those that
results in more properties lost. We also separated the outstanding events
(identifying outliers) to emphasize on the less publicized but more common
ones.

## Data Processing

For this study we used the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. It records major weather
events since 1950 including location, duration and estimates in
property damages and fatalities and injuries.

The data comes in a cvs compressed file.
Estimates of damages are coded in thousands, millions and billions
dollars. We transformed them to dollars units.

```{r}
library(ggplot2)
Sys.setlocale("LC_TIME", "C")
stormdata <- read.csv(bzfile('repdata-data-StormData.csv.bz2'),
                      stringsAsFactors = F)

todollars <- function(ammount, expfactor) {
    if (expfactor == "K") { ammount * 1000 }
    else if (expfactor == "M") { ammount * 1000000 }
    else if (expfactor == "B") { ammount * 1000000000 }
    else { ammount }
}

storms <- data.frame(as.Date(stormdata$BGN_DATE, '%m/%d/%Y'),
                     stormdata$STATE,
                     as.factor(stormdata$EVTYPE),
                     stormdata$FATALITIES,
                     stormdata$INJURIES,
                     mapply(todollars, stormdata$PROPDMG, stormdata$PROPDMGEXP),
                     mapply(todollars, stormdata$CROPDMG, stormdata$CROPDMGEXP),
                     stringsAsFactors=F
)

names(storms) <- c("date","state","evtype",
                   "fatalities","injuries","propdmg","cropdmg")
```

As stated in http://www.ncdc.noaa.gov/stormevents/details.jsp the code for
events is more detailed since January 1996. The last recorded event in
this data file is November 2011. We filtered the data and, from now on,
we only processed events from December 1995 so we had 6 complete years.
We also didn't differentiate between property and crop (agriculture)
damages and considered both of them together.

```{r}
storms96 <- storms[which(as.numeric(format(storms$date, "%Y%m")) >= 199512),]
storms96$damages <- storms96$propdmg + storms96$cropdmg

```


## Results

### Human Life Costs

We grouped the event types by the casualties they cause.

```{r}
fatlpertype <- aggregate(fatalities ~ evtype,
                    data=storms96[storms96$fatalities > 0,],
                    FUN=sum)

injrpertype <- aggregate(injuries ~ evtype,
                         data=storms96[storms96$injuries > 0,],
                         FUN=sum)
```                         

Most the bad weather events didn't cause any victims.

```{r}
1 - round(sum(storms96$fatalities == 0 & storms96$fatalities == 0)
      / nrow(storms96) * 100, 1)
```

But some of them were breathless and hard to forget like the Joplin
tornado and the heatwaves in the Midwest.

```{r}
storms96[storms96$fatalities >= 50, c(3,1,2,4,5)]
```

The 5 worst events presented in descending order according
the fatalities per year they cause are:

```{r}
worstfatl <- order(fatlpertype$fatalities, decreasing=T)
fatlpertype$fatalitiesperyear <- round(fatlpertype$fatalities / 6)
head(fatlpertype[worstfatl, c("evtype","fatalitiesperyear")], n=5)
```

The same table ordered by injuries:

```{r}
worstinjr <- order(injrpertype$injuries, decreasing=T)
injrpertype$injuriesperyear <- round(injrpertype$injuries / 6)
head(injrpertype[worstinjr, c("evtype","injuriesperyear")], n=5)
```

### Damages

Regarding damages we calculated the total damages per year in US dollars
for each bad weather event type.

```{r}
dmgspertype <- aggregate(damages ~ evtype,
                         data=storms96[storms96$damages > 0,],
                         FUN=sum)
worstdmgs <- order(dmgspertype$damages, decreasing=T)
# millions per year
dmgspertype$damagesperyear <- round(dmgspertype$damages / 6000000)
head(dmgspertype[worstdmgs, c("evtype","damagesperyear")], n=5)
```

Filtering outliers shows that storms and bad weather causes
great lost in properties and agriculture throughout the year
summing up high ammount of liabilities.

```{r}
dmgspermonth <- aggregate(damages ~ evtype + format(date, "%m %Y"),
                          data=storms96, FUN=sum)
names(dmgspermonth)[2] <- "when"
worstevents <- head(dmgspertype[worstdmgs, c("evtype")], n=5)
dmgspermonthworst5 <- dmgspermonth[dmgspermonth$evtype %in% worstevents,]
dmgspermonthworst5$dwhen <- as.Date(paste("1 ", dmgspermonthworst5$when),
                                    format="%d %m %Y")
outl <- boxplot(dmgspermonthworst5$damages)$out
ggplot(subset(dmgspermonthworst5, ! damages %in% outl),
       aes(x=dwhen, y=damages/1000000)) +
    geom_line(aes(group=1)) +
    facet_grid(evtype~.) +
    labs(x="Date of Event", y="Damages (millions of US$)",
         title="Damages per type since December 1995.")
```
_Figure 1: Damages since December 1995 in millions of US$ without outliers.
It's only shown the 5 worst bad weather event types._

## Conclusions

BAD WEATHER.


```{r}
1+1
```

